---
title: "MATH 532 Project Presentation"
subtitle: "Mpox Surveillance and Reporting in Nigeria: Insights from Count Data Modeling"
author: "Aliu A. Adebiyi"
date: "`r Sys.Date()`"
output: beamer_presentation
theme: "Boadilla"
colortheme: "whale"
fonttheme: "serif"
fontsize: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
#libraries
library(faraway)
library(ggplot2)
library(tidyverse)
library(readxl)
```

## Introduction

- Monkeypox is a zoonotic disease caused by infection with the Monkeypox Virus (MPXV), an orthopoxvirus with symptoms similar but less severe than smallpox (Precious et al., 2023).

- Its burden is more pronounced in developing countries, making the outbreak a global concern (Precious et al., 2023).

- In Nigeria, preparedness and timely response are vital, hence the recent effort by the Nigeria Centre for Disease Control (NCDC) to collate reported cases across states.

- This project analyzes Mpox data (2017–2024) using count regression models to identify factors influencing reported cases across Nigerian states.

## Study Design and Data Sources
- This is an ecological study of Mpox cases reported in Nigeria between 2017 and 2024.

- Data on reported cases were obtained from the NCDC (NCDC, 2024).

- Explanatory variables—population density, GDP per capita, literacy, temperature, and region—were sourced from the Nigerian Bureau of Statistics (NBS, 2021).

- Variable selection was based on prior studies on COVID-19 distribution in Nigeria (Daniel & Adejumo, 2021).

## Count Regression Modelling
Let \( Y_i \) be the number of reported cases. Since the response variable is count data, we use a **Poisson regression model**(Faraway, 2016):

\[
Y_i \sim \text{Poisson}(\mu_i), \quad \text{where} \quad E(Y_i) = \mu_i, \quad \text{Var}(Y_i) = \mu_i
\]

A **log link function** relates the mean to the predictors:

\[
\log(\mu_i) = \mathbf{X}_i^T \boldsymbol{\beta}
\quad \Rightarrow \quad
\mu_i = \exp(\mathbf{X}_i^T \boldsymbol{\beta})
\]

Here, \( \mathbf{X}_i \) is a vector of predictors and \( \boldsymbol{\beta} \) is the vector of coefficients.

## Procedure for Count Modelling Analysis
- Plotted the response variable against one or more predictors.

- Fitted a Poisson model using all predictors.

- Checked deviance and residual plots for model fit.

- Evaluated the dispersion parameter to detect overdispersion.

- Fitted a Negative Binomial model to handle overdispersion and checked diagnostics.

- Compared the full model with the null (intercept-only) model.

- Used stepwise selection to identify significant predictors and fitted a reduced model.

- Interpreted the final model and drew conclusions.

## Data Exploration
```{r}
#loading the data
mpox <- read_xlsx("mpox.xlsx", sheet = "mpox_total_states")
```

```{r, echo=TRUE}
head(mpox)
```

```{r, echo=TRUE}
dim(mpox)
```

## Data Exploration

```{r}
# Summary Statistics by Region
region_summary <- mpox %>%
  group_by(region) %>%
  summarise(
    total_cases = sum(cases, na.rm = TRUE),
    avg_reported_cases = mean(cases, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(desc(avg_reported_cases))

ggplot(region_summary, aes(x = reorder(region, -avg_reported_cases), 
                         y = avg_reported_cases, fill = avg_reported_cases)) +
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "lightblue", high = "blue", name = "Cases/million") +
  labs(title = "Average Reported Mpox Cases by Region", 
       x="Region",y="Cases reported",
       caption="Data source: Nigeria CDC and Nigeria Bureau of Statistics") +
  theme_bw()
```

## Analysis - Scatter Plots
```{r}
#Scatter plot
par(mfrow = c(2, 2))
plot(mpox$pop_density, mpox$cases,
     xlab = "Population Density (people/km²)", 
     ylab = "Reported Mpox Cases",
     main = "Cases vs. Population Density",
     pch = 19, col = "steelblue")
abline(lm(cases ~ pop_density, data = mpox), col = "red", lwd = 2)

plot(mpox$gdp_per_capital, mpox$cases,
     xlab = "GDP per Capita ('000$)", ylab=" ",
     main = "Cases vs. GDP per Capita",
     pch = 19, col = "darkgreen")
abline(lm(cases ~ gdp_per_capital, data = mpox), col = "red", lwd = 2)

plot(mpox$literacy, mpox$cases,
     xlab = "Literacy Rate (%)", 
     ylab = "Reported Mpox Cases",
     main = "Cases vs. Literacy",
     pch = 19, col = "purple")
abline(lm(cases ~ literacy, data = mpox), col = "red", lwd = 2)

plot(mpox$avg_temp, mpox$cases,
     xlab = "Average Temperature (°C)", ylab = " ",
     main = "Cases vs. Temperature",
     pch = 19, col = "orange")
abline(lm(cases ~ avg_temp, data = mpox), col = "red", lwd = 2)
```

## Analysis - Poisson Model.
```{r}
mpox$region<-factor(mpox$region)
mpox$region<-relevel(mpox$region, ref = "North") #making north the reference category
```

```{r cars, echo=TRUE}
pois_mod <- glm(cases~.-state, data=mpox, family="poisson")
sumary(pois_mod)
```

## Analysis - Diagnostic
**Plots of Residuals, Poisson Model**
```{r}
resid_dev <- rstandard(pois_mod) #Standardized deviance residuals
qqnorm(resid_dev, main = "QQ Plot of Deviance Residuals") # QQ plot
qqline(resid_dev, col = "red")
```

## Analysis - Diagnostic

**Over-dispersion Parameter**
```{r, echo=TRUE}
dp = pois_mod$deviance/pois_mod$df.residual
dp
```

This necessitate the use of count regression modeling that account for over dispersion parameters as an alternative.

## Analysis - Negative Binomial Regression
```{r}
library(MASS)
```

```{r, echo=TRUE}
nb_mod <-glm.nb(cases ~ . -state, data = mpox)
nb_mod
```
## Analysis - AIC criterion for significant predictors
```{r, echo=TRUE}
step(nb_mod, trace = 0)
```
## Analysis - Reduced Model
```{r, echo=TRUE}
reduced_mod <- glm.nb(cases~gdp_per_capital+literacy -state, data = mpox)
summary(reduced_mod)
```
## Analysis - Diagnostic
```{r}
resid_nb_r <- rstandard(reduced_mod) #Standardized deviance residuals
qqnorm(resid_nb_r, main = "QQ Plot of Deviance Residuals") # QQ plot
qqline(resid_nb_r, col = "red")
```

## Result and Interpretation
```{r }
library(effects)
eff <- allEffects(reduced_mod)
print(plot(eff))
```

## Result and Interpretation 
```{r, echo=TRUE}
exp(coef(reduced_mod))
```

The coefficient of GDP per capita gives $exp(0.747) = 2.11$ indicating that for every unit increase in gdp_per_capital, the expected number of Mpox cases increases by a factor of 2.11 holding literacy constant. Also, Literacy has a coefficient of $exp(0.019) = 1.02$ meaning that for every 1% increase in literacy rate, the expected number of Mpox cases increases by about 2%, holding GDP constant. \
\
The implication of this result is that higher reported cases in wealthier and more literate regions may reflect better surveillance, health infrastructure, and awareness, rather than higher actual transmission.

## Conclusion 
- A Negative Binomial model was used to examine factors influencing reported Mpox cases in Nigeria.

- After adjusting for over-dispersion, GDP per capita and literacy rate emerged as key predictors.

- Regions with higher GDP per capita were over 2 times more likely to report Mpox cases, suggesting stronger surveillance and reporting systems.

- Literacy showed a modest positive effect, indicating the role of awareness and information access in disease detection.

- Results underline the importance of socioeconomic and educational factors in public health surveillance across regions.

## References

- Daniel, O. J., & Adejumo, O. A. (2021). *Spatial distribution of COVID-19 in Nigeria*. West African Journal of Medicine, 38(8), 732–737.

- Faraway, J. J. (2016). *Extending the Linear Model with R: Generalized Linear, Mixed Effects and Nonparametric Regression Models* (2nd ed.). Chapman and Hall/CRC.

- Nigerian Centre for Disease Control (NCDC). (2024). *Update on Mpox (MPX) in Nigeria*. Retrieved from [www.ncdc.gov.ng](https://www.ncdc.gov.ng)

- Nigerian Data Portal. (2021). *State population 2006*. Retrieved from [https://nigeria.opendataforafrica.org/ifpbxbd](https://nigeria.opendataforafrica.org/ifpbxbd)

- Precious, N. D., Agboola, P., Oluwatimilehin, O., Olakunle, O. K., Olaniyi, P., Adiatu, A. I., Olusogo, A. P., Obiwulu, D. J., Adeola, O. A., Ebubechukwu, E. S., Oluwakayode, A. M., Akano, O. S., & Kolawole, Q. O. (2023). Re-emergence of monkeypox virus outbreak in Nigeria: Epidemic preparedness and response (Review-Commentary). *Annals of Medicine and Surgery*, 85(8), 3990–3996. https://doi.org/10.1097/MS9.0000000000001069}


## Final Slide

\begin{center}
\Huge \textbf{THANK YOU}
\end{center}

\
Download the code and the complete Rmarkdown file from [\textcolor{blue}{my github}](https://github.com/Alisonia?tab=repositories) 
